Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies

### 2018-09-20 legacy.tex

While we were unable to reproduce the heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
We also found that this batch effect led to spurious results in a additional recent publications as a consequence of biased imputation.

The results section is organized as follows.
First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
Finally, we investigate how these SNPs have impacted modern genomics analyses. 
We report that major imputation servers impute the dubious SNPs. 
We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.

@@

Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a}.
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants.

@@

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
In this case, we are using simple logistic regression looking for loci where the genotypes are predictable based on the data quality scores.
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

	\subsection{Identifying Suspicious SNPs in the 1kGP}
The results of the \textit{reverse} GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
By 2011, older sequencing technologies were phased out, and method development solidify, which coincides with the sequencing quality levelling off.
It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
However, the $Q$ per individual appeared to be the best metric to identify spurious sites.

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan}.

@@

@@

\subsection{Suspicious SNPs impact modern genomics analyses}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
We found that all of the SNPs we identified as being associated with $Q$ for chromosomes 1 and 2 were imputed. 
This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants have not been removed.

>>>>

While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
We also found that this batch effect led to spurious imputation results which found their way in recent publications.

The results section is organized as follows.
We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1000 genomes project. 
Next, we expand our analysis to the other populations in the 1kGP and identify lists of suspicious SNPs.
Finally, we investigate how these SNPs have impacted modern genomics analyses.

@@

Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a} \sgcomment{That would be 2017, no?}.
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants. \sgcomment{this sentence needs work}

@@

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use linear regression of a phenotype on a genotype. 
Here we use logistic regression of the genotypes on a quantitative measure of quality (i.e., the genotype is the dependent variable).
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

	\subsection{Identifying Suspicious SNPs in the 1kGP}
The results of the \textit{reverse} GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.

By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases ($Q$) per individual because it best correlated with mismatched sites across Japanese cohorts.

@@

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan}.

@@

\subsection{Suspicious SNPs impact modern genomics analyses}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.

\sgcomment{In order to -> To, in general} In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
We found that all of the \sgomment{"all of the" -> "all the" } SNPs we identified as being associated with $Q$ were imputed back in the samples. 
This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
